Missing The Point

### 2020-10-20 Manuscript.Rmd
Add iterations to case study

Iterative imputation is a popular tool to accommodate the ubiquitous problem of missing data. While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence. Our study provides insight into identifying non-convergence in iterative imputation algorithms since there is no consensus on how to evaluate the convergence properties currently. We found that--in the cases considered--inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods. We conclude that it never hurts to iterate longer, but such calculations hardly bring added value.

@@

Missing data pose a ubiquitous threat to anyone who aims to obtain unbiased, confidence-valid statistical inferences. A popular technique to accommodate missing data is to 'impute' (i.e., fill in) any missing values in an incomplete dataset. Imputation procedures like 'Multiple Imputation by Chained Equations' (MICE) have proven to be powerful tools to draw valid inference under many missing data circumstances (Rubin, 1987; van Buuren, 2018). To obtain imputations, most imputation software packages rely on iterative algorithms.

With iterative imputation, the validity of the inference depends on the state-space of the algorithm at the final iteration. This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained using the imputed data? These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied (Van Buuren, 2018, 6.5.2). While there is no scientific consensus on how to evaluate the convergence of imputation algorithms (Zhu & Raghunathan, 2015; Takahashi, 2017), the current practice is to visually inspect imputations for signs of non-convergence.

@@

We investigate non-convergence in iterative imputation through model-based simulation in R (version 4.0.2; R Core Team 2020). We provide a summary of the simulation set-up in Algorithm 1; the complete script and technical details are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).

@@

We aim to assess the impact of non-convergence on the validity of scientific estimates obtained using `mice` [@mice]. Inferential validity is reached when estimates are both  unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$). To induce non-convergence we terminate the iterative algorithm at different imputation chain lengths ($n_{it} = 1, 2, ..., 50$). We differentiate between nine different missingness scenarios: three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??].

@@

We use eight different diagnostic methods to identify non-convergence: a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters of interest---chain means, chain variances, a scientific estimate, and the novel parameter we propose, $\lambda$.

@@

The following plots provide an overview of the results for each method and each performance measure, contrasted to the number of iteration in the imputation algorithm. Within the plots, we split the results according to the missingness conditions [missingness mechanisms as line types, and proportion of incomplete cases as colors].

@@

In this plot, our diagnostic method is the amount of autocorrelation in the chain means. We can interpret this as the extent to which the average imputed value trends across iterations. We see that autocorrelation in the chain means rapidly decreases until $it \geq 6$. This means that there is some initial trending within chains, but stationarity does not improve substantively after the first few iterations. These results hold irrespective of the missingness condition.

@@

Autocorrelation in the chain variances shows something similar. The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally when $it \geq 10$. We do not observe a systematic effect of the missingness conditions here either.

@@

There is more autocorrelation in the scientific estimates than in the univariate parameters (chain means and chain variances). We observe the highest autocorrelations in conditions where 75% of cases are incomplete. Overall, the autocorrelations reach a plateau when $it \geq 20$ to $30$. There is no clear effect of the missingness mechanisms.

@@

We observe that $\widehat{R}$ in the chain means generally decreases as a function of the number of iterations. An exception to this observation is the initial increase when $3 < it < 5$. According to these results, the mixing between chain means generally improves until $it \geq 30$ to $40$. There is no apparent differentiation between the missingness conditions.

@@

With the scientific estimate as parameter we observe very similar $\widehat{R}$-values again. We do, however, see some differentiation between missingness conditions. Conditions where 75% of the cases are incomplete show more extreme non-mixing. The overall trend remains the same: about 30 iterations are required before mixing stops improving substantially.

@@

$\widehat{R}$-values in the chains of the novel parameter show a trend similar to the chain means and chain variances. [add something about conditions??]

@@

```{r echo=FALSE}
p <- simulation_plot(v="rsq", name = "Bias in coefficient of determination") +
  list(geom_hline(yintercept = 0, color = "grey"))
p # %>% plotly::ggplotly()
```
Equivalent to the bias in the regression estimate, the bias in the coefficient of determination tapers off within a couple of iterations. We observe stable estimates in all conditions when $it \geq 6$ [interpret the over-estimation in MNAR+75% condition?].

@@

@@

```{r echo=FALSE}
p <- casestudy_plot(v="r.hat.max.sigma.hgt", name = "Rhat chain variance")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
p # %>% plotly::ggplotly()
```

```{r echo=FALSE}
p <- casestudy_plot(v="r.hat.max.qhat", name = "Rhat regression estimate")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
p # %>% plotly::ggplotly()
```

```{r echo=FALSE}
p <- casestudy_plot(v="r.hat.max.lambda", name = "Rhat lambda")+list(geom_hline(yintercept = 1.2, color = "grey"), scale_y_continuous(limits = c(1,2)))
p # %>% plotly::ggplotly()
```

>>>>

Iterative imputation is a popular tool to accommodate the ubiquitous problem of missing data. While it is widely accepted that this technique can yield valid inferences, these inferences all rely on algorithmic convergence. Since there is no consensus on how to evaluate the convergence properties iterative imputation algorithms, our study provides insight into identifying non-convergence. We found that--in the cases considered--inferential validity was achieved after five to ten iterations, much earlier than indicated by diagnostic methods. We conclude that it never hurts to iterate longer, but such calculations hardly bring added value.

@@

Missing data pose a ubiquitous threat to anyone who aims to obtain unbiased, confidence-valid statistical inferences. A popular technique to accommodate missing data is to 'impute' (i.e., fill in) any missing values in an incomplete dataset. Imputation procedures like 'Multiple Imputation by Chained Equations' (MICE) have proven to be powerful tools to draw valid inference under many missing data circumstances [@rubin87; @buur18]. To obtain imputations, most imputation software packages rely on iterative algorithms.

With iterative imputation, the validity of the inference depends on the state-space of the algorithm at the final iteration. This introduces a potential threat to the validity of the imputations: What if the algorithm has not converged? Are the imputations then to be trusted? And can we rely on the inference obtained using the imputed data? These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied (Van Buuren, 2018, 6.5.2). While there is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @taka17], the current practice is to visually inspect imputations for signs of non-convergence. 
<!-- (Zhu & Raghunathan, 2015; Takahashi, 2017) -->

@@

We investigate non-convergence in iterative imputation through model-based simulation in R (version 4.0.3; R Core Team 2020). We provide a summary of the simulation set-up in Algorithm 1; the complete script and technical details are available from [github.com/hanneoberman/MissingThePoint](https://github.com/hanneoberman/MissingThePoint).

@@

With this simulation, we assess the impact of non-convergence on the validity of scientific estimates obtained using `mice` [@mice]. Inferential validity is reached when estimates are both unbiased and have nominal coverage across simulation repetitions ($n_{sim} = 1000$). To induce non-convergence, we terminate the iterative algorithm at different imputation chain lengths ($n_{it} = 1, 2, ..., 50$). We differentiate between nine different missingness scenarios: three missingness mechanisms versus three proportions of incomplete cases [remove here if already defined above/below??].

@@

We use eight different diagnostic methods to identify non-convergence: a combination of two non-convergence identifiers---autocorrelation and $\widehat{R}$---and four parameters of interest---chain means, chain variances, a scientific estimate, and the novel parameter that we propose, $\lambda$.

@@

The following figures provide an overview of the results for each method and each performance measure, contrasted to the number of iterations in the imputation algorithm. Within the figures, we split the results according to the missingness conditions [missingness mechanisms as line types, and proportion of incomplete cases as colors].

@@

In this figure, we see autocorrelation in the chain means as the diagnostic method. We can interpret this as the extent to which the average imputed value trends across iterations [unnecessary??]. We observe that autocorrelation in the chain means rapidly decreases until $it \geq 6$. This means that there is some initial trending within chains, but stationarity does not improve substantively after the first few iterations. These results hold irrespective of the missingness condition.

@@

Autocorrelation in the chain variances show us something similar. The number of iterations that is required to reach non-improving autocorrelations is somewhat more ambiguous than for chain means, but generally around $it \geq 10$. We do not observe a systematic difference between missingness conditions here either.

@@

There is more autocorrelation in the scientific estimates than in the univariate parameters (chain means and chain variances). We observe the highest autocorrelations in conditions where 75% of cases are incomplete. Overall, the autocorrelations reach a plateau when $it \geq 20$ to $30$. There is no clear effect of the missingness mechanisms [unnecessary?].

@@

We observe that $\widehat{R}$-values of the chain means generally decreases as a function of the number of iterations. An exception to this observation is the initial increase when $3 \leq it \leq 5$ [interpret?? due to initialization or is there really more mixing initially??]. After the first couple of iterations, the mixing between chain means generally improves until $it \geq 30$ to $40$. There is no apparent differentiation between the missingness conditions.

@@

With the scientific estimate as parameter we observe very similar $\widehat{R}$-values again. We do, however, see some differences between missingness conditions. Conditions where 75% of the cases are incomplete show more extreme non-mixing. The overall trend remains the same: about 30 iterations are required before mixing stops improving substantially.

@@

$\widehat{R}$-values of the novel parameter show a trend similar to the chain means and chain variances. [add something about conditions??]

@@

```{r echo=FALSE}
p <- simulation_plot(v="rsq", name = "Bias in coefficient of determination") +
  list(geom_hline(yintercept = 0, color = "grey"))
p # %>% plotly::ggplotly()
```

Equivalent to the bias in the regression estimate, the bias in the coefficient of determination tapers off within a couple of iterations. We observe stable estimates in all conditions when $it \geq 6$ [interpret the over-estimation in MNAR+75% condition?].

@@

From the traceplot of the chain means it seems that mixing improves up-to 20 iterations, and there is some trending up-to 20 iterations as well. The chain variances are more or less stationary after 5 iterations, and mixing requires about 20 iterations.

@@

```{r echo=FALSE}
p <- casestudy_plot(v="r.hat.max.mu.hgt", name = "Rhat chain means") + 
  list(
    geom_hline(yintercept = 1.2, color = "grey")#, 
    #scale_y_continuous(limits = c(1,5))
    )
p # %>% plotly::ggplotly()
```

```{r echo=FALSE}
p <-
  casestudy_plot(v = "r.hat.max.qhat", name = "Rhat regression estimate") +
  list(geom_hline(yintercept = 1.2, color = "grey")#,
       #scale_y_continuous(limits = c(1, 2.5))
       )
p # %>% plotly::ggplotly()
```

```{r echo=FALSE}
p <- casestudy_plot(v = "r.hat.max.lambda", name = "Rhat lambda") +
  list(geom_hline(yintercept = 1.2, color = "grey")#,
       #scale_y_continuous(limits = c(1, 2.5))
       )
p # %>% plotly::ggplotly()
```
